\documentclass[a4paper,11pt,oneside]{article}
\usepackage{url}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{comment}

%%%% Packages: %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

% \usepackage[english]{babel} 
\usepackage[english]{babel} 

\usepackage{cite}

\usepackage{geometry}
\geometry{textwidth=16cm,textheight=26cm}

%%%% Preamble: %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{D1. Topic of interest: Design of complex CI pipelines
}

\author{
    Brice Delcroix
    \and Timothé Piscaglia
    \and Pierre-Louis Lefèvre}
    
\date{
    Vint Cerf, INFO B302 - IDS (2025 - 2026)
}
    
%%%% Document: %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

% ---------------------------------------
\section{Topic of interest}
% ---------------------------------------

Our topic addresses the design, architecture, and management of complex Continuous Integration (CI) pipelines. We will focus on the subject of identifying intricate pipeline requirements, modeling efficient workflows, and handling diverse and scalable CI configurations.

% ---------------------------------------
\section{Short description}
% ---------------------------------------

Continuous Integration (CI) is a fundamental practice in software engineering, triggered by specific events such as code pushes or pull requests. As projects are added, pipelines can become very complex and difficult to maintain. The design of complex pipelines requires a thorough understanding of project-specific needs, including varying environments, interdependent modules, and diverse testing frameworks. The systematic modeling is an essential aspect of CI pipelines, which allows development teams to visualize, analyze, and optimize workflows prior to full-scale implementation. \\

Furthermore, managing CI configurations efficiently often means treating configuration as code ensures that pipelines remain scalable, reproducible, and resilient to human error over time. Mastering the design of complex CI pipelines leads to faster feedback loops, more reliable software releases, and a streamlined development process across large engineering teams.
\end{document}
